---
title: "Data mockups"
author: "Lan Kelly"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:   
  word_document: 
  reference_docx: rmarkdown-styles-SAP.docx
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE, 
error=FALSE, comment='', dpi=400)
options(knitr.kable.NA = '')
options(width=1000) # Wide pages
options(scipen=999) # avoid scientific presentation
source = 'real_data'
#source = 'dummy_data'
source('1_ecmo_ich_data_prep.R') # prepares the data and runs exclusions
library(visdat)
library(summarytools)
library(pander)
library(naniar)
library(arsenal)
```

\newpage
# Comparison of V-V ECMO patients with and without stroke
The following tables are based on the document Data Mock Ups_V3.docx.

```{r, results='asis'}
nstroke_before = nrow(ecmo_patients %>%
  filter(stroke_after_ECMO == "No"))

ecmo_patients <- ecmo_patients %>% filter(ecmo_type == "Venous-Venous")

library(tableone)
library(labelled)
vars <- c('age','sex','ethnicity','country','bmi','apache_ii','sofa','pregnant',
          "comorbidity_obesity" ,'comorbidity_immuno','comorbidity_hypertension',
          'comorbidity_diabetes',"comorbidity_smoking","comorbidity_chronic_cardiac_disease",
          'comorbidity_chronic_kidney_disease','comorbidity_chronic_alcohol_abuse_eot',
          "comorbidity_malignant_neoplasm",'comorbidity_severe_liver_disease')


tabdata <- ecmo_patients %>%
  mutate(complication_stroke = ifelse(is.na(complication_stroke), 'Missing', 
                                      as.character(complication_stroke))) %>%
  filter(stroke_after_ECMO != "No" | is.na(stroke_after_ECMO)) %>%
  select(all_of(vars), complication_stroke) %>%
  set_variable_labels(age =  "Age",sex="Sex",ethnicity="Ethnicity",country='Country',
                      bmi='BMI',apache_ii='APACHE II',sofa='SOFA',
                      pregnant='Pregnant',comorbidity_obesity = "Comorbid obesity" ,
                      comorbidity_immuno='Comorbid immunosuppression',
                      comorbidity_hypertension='Comorbid hypertension',
                      comorbidity_diabetes='Comorbid diabetes',
                      comorbidity_smoking="Comorbid smoking",
                      comorbidity_chronic_cardiac_disease="Comorbid chronic cardiac disease",
                      comorbidity_chronic_kidney_disease='Comorbid CKD',
                      comorbidity_chronic_alcohol_abuse_eot='Comorbid alcohol abuse',
                      comorbidity_malignant_neoplasm="Comorbid malignant neoplasm",
                      comorbidity_severe_liver_disease='Comorbid severe liver disease',
                      complication_stroke="Stroke")
```

\newpage
For the `r nrow(tabdata)` V-V ECMO patients, excluding the `r nstroke_before` patients who had a stroke before ECMO. 

```{r, results='asis'}

#var_label(tabdata)

#still need RR at time of intubation, time from intubation to ECMO
cat_vars = c('sex','ethnicity','country','pregnant',
          "comorbidity_obesity" ,'comorbidity_immuno','comorbidity_hypertension',
          'comorbidity_diabetes',"comorbidity_smoking","comorbidity_chronic_cardiac_disease",
          'comorbidity_chronic_kidney_disease','comorbidity_chronic_alcohol_abuse_eot',
          "comorbidity_malignant_neoplasm",'comorbidity_severe_liver_disease')

skewvars <- c('age','bmi','apache_ii','sofa')


## Create a TableOne object
tab1 <- CreateTableOne(vars = vars, data = tabdata, factorVars = cat_vars,
                       includeNA = T,
                       strata="complication_stroke")


kableone(print(tab1,test=F, varLabels=T,pDigits=2, missing = F,
               caption = "Patient Characteristics")) #nonnormal = skewvars, 

```

\newpage
### Patients pre V-V ECMO respiratory support characteristics

Pre-ECMO characteristics for the `r nrow(tabdata)` V-V ECMO patients, excluding the `r nstroke_before` patients who had a stroke before ECMO.

```{r, results='asis'}

#var_label(tabdata)

#still need static compliance


ecmo_vars <-c("ecmo_highest_fi_o2_6hr_before","ecmo_highest_peep_6hr_before",
         "ecmo_worst_pa_o2_6hr_before","ecmo_worst_pa_co2_6hr_before",
         "ecmo_prone_before" ,"ecmo_neuromuscolar_blockage_before",
         "days_icu",'days_hosp', 'days_ecmo','days_vent',
          'days_sympt_hosp','day',"outcome")
cat_vars = c('outcome')

skewvars <- c("ecmo_highest_fi_o2_6hr_before","ecmo_highest_peep_6hr_before",
         "ecmo_worst_pa_o2_6hr_before","ecmo_worst_pa_co2_6hr_before",
         "ecmo_prone_before" ,"ecmo_neuromuscolar_blockage_before",
         "days_icu",'days_hosp', 'days_ecmo','days_vent',
          'days_sympt_hosp','day')


tabdata <- ecmo_patients %>%
  mutate(complication_stroke = ifelse(is.na(complication_stroke), 'Missing', 
                                      as.character(complication_stroke))) %>%
  filter(stroke_after_ECMO != "No" | is.na(stroke_after_ECMO)) %>%
  select(all_of(ecmo_vars), complication_stroke) %>%
  set_variable_labels(ecmo_highest_fi_o2_6hr_before = "Pre-ECMO FiO2",
                      ecmo_highest_peep_6hr_before = "Pre-ECMO PEEP",
                      ecmo_worst_pa_o2_6hr_before = "Pre-ECMO PaO2",
                      ecmo_worst_pa_co2_6hr_before = "Pre-ECMO PaCO2",
                      ecmo_prone_before = "Pre-ECMO prone positioning",
                      ecmo_neuromuscolar_blockage_before = "Pre-ECMO neuromuscular blockade",
                      days_icu = "Days in ICU",
                      days_ecmo='Length of the ECMO run',
                      days_hosp='Hospital length of stay',
                      days_vent="Days of mechanical ventilation",
                      days_sympt_hosp="Days from first symptoms to hospitalisation",
                      day='Day put on ECMO',
                      outcome='Outcome',
                      complication_stroke="Stroke")


## Create a TableOne object
tab2 <- CreateTableOne(vars = ecmo_vars, data = tabdata, factorVars = cat_vars,
                       includeNA = T,
                       strata="complication_stroke")


kableone(print(tab2,test=F,varLabels=T,pDigits=2,  missing = F,
               caption = "Patients pre-ECMO respiratory support characteristics")) #nonnormal = skewvars,

```
