---
title: "Plots of lab values"
author: "Lan Kelly"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:   
  word_document: 
  #reference_docx: rmarkdown-styles-SAP.docx
  reference_docx: WordStylesReference01.docx
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE, 
error=FALSE, comment='', dpi=400)
options(knitr.kable.NA = '')
options(width=1000) # Wide pages
options(scipen=999) # avoid scientific presentation
# options(fig.width=10)
# options(out.height=10)
# opts_knit$set(global.par = TRUE)


```

```{r, include = F}
knitr::read_chunk("2_univariate_survival.R")
```

```{r get_data}

```


\newpage

# Lab values after ECMO initiation



```{r, include=FALSE}
# data to plot, must have dates

library(ggplot2)
library(dplyr)
fup = 21

ecmo_daily <- ecmo_daily %>% 
  filter(ecmo_type == "Venous-Venous",
         !is.na(ecmo), !is.na(eotd_anticoagulants)) %>%
  mutate(il_6=ifelse(il_6>10000, NA_real_, il_6),
         d_dimer=ifelse(d_dimer>70000, NA_real_, d_dimer),
         aptt=ifelse(aptt>200, NA_real_, aptt)
         )

to_plot = filter(ecmo_daily, 
                 !is.na(date_ecmo),
                 !is.na(date_daily),
                 !is.na(complication_stroke)) %>%
  mutate(day = as.numeric(date_daily - date_ecmo)) %>%
  filter(day <= fup & day >= 0) # exclude long stays for the plot


to_plot = mutate(to_plot, facet = paste("Stroke = ", complication_stroke, sep='')) # nice facet

z = qnorm(0.975)


boxplot_over_time <- function(var, ylabel) {
  
  p <- to_plot %>%
    ggplot(aes(x=factor(day), y={{var}}, col=factor(facet)))+ 
  geom_boxplot()+
  xlab('Day after ECMO initiation')+
  ylab(ylabel)+
  scale_color_manual('Stroke', values=c('darkorange1','cyan4'))+ #'Any ECMO',,'cyan4'
  facet_wrap(~facet, scales='fixed', nrow=2) +
  g.theme +
  theme(legend.position = "none")
  
  return(p)
  
}

means_over_time <- function(var, ylabel, log=T) {
  
  if (log) {
    to_stat = group_by(to_plot, day, facet) %>%
      summarise(mean = mean(log({{var}}), na.rm = TRUE),
                se = sem(log({{var}}), na.rm = TRUE)) %>%  
      ungroup() %>%
      mutate(lower = exp(mean - (z*se)),
             upper = exp(mean + (z*se)),
             mean=exp(mean)
         #day = ifelse(any_ecmo=='Yes', day+0.2, day)
         ) # jitter x slightly to avoid overlap
    
  } else {
    to_stat = group_by(to_plot, day, facet) %>%
      summarise(mean = mean({{var}}, na.rm = TRUE),
            se = sem({{var}}, na.rm = TRUE)) %>%  #sem
      ungroup() %>%
      mutate(lower = mean - (z*se),
         upper = mean + (z*se)
         ) 
    
  }
  
p <-to_stat %>%
  ggplot(aes(x=day, y=mean, ymin=lower, ymax=upper, col=factor(facet)))+
  geom_point()+
  geom_errorbar(width=0)+
  xlab('Day after ECMO initiation')+
  ylab(ylabel)+
  facet_wrap(~facet, scales='fixed', nrow=2) +
  scale_color_manual('Stroke', values=c('darkorange1','cyan4'))+ #'Any ECMO', ,'cyan4'
  g.theme +
  theme(legend.position = "none")

return(p)
  
}

```

This document plots the lab value trajectories for the first `r fup` days after the start of ECMO support, for `r length(unique(ecmo_daily$pin))` patients who did or did not experience a stroke. 


\newpage
## PaO2

##### Boxplot over time

```{r, fig.width=8}

tplot = boxplot_over_time(pa_o2, 'PaO2')
tplot
```

##### Plot of means and 95% confidence intervals for the means over time

```{r, fig.width=8}

tplot = means_over_time(pa_o2, 'PaO2')
tplot
```


\newpage
## PaCO2


##### Boxplot over time

```{r, fig.width=8}

tplot = boxplot_over_time(pa_co2, 'PaCO2')
tplot
```

##### Plot of means and 95% confidence intervals for the means over time

```{r, fig.width=8}

tplot = means_over_time(pa_co2, 'PaCO2')
tplot
```



\newpage
## PaO2/FiO2



##### Boxplot over time

```{r, fig.width=8}

tplot = boxplot_over_time(pa_o2_fi_o2, 'PaO2/FiO2')
tplot
```

##### Plot of means and 95% confidence intervals for the means over time

```{r, fig.width=8}

tplot = means_over_time(pa_o2_fi_o2, 'PaO2/FiO2')
tplot
```






\newpage
## Platelet count

##### Boxplot over time

```{r, fig.width=8}


tplot = boxplot_over_time(platelet_count, 'Platelet count')
tplot
```

##### Plot of means and 95% confidence intervals for the means over time

```{r, fig.width=8}


tplot = means_over_time(platelet_count, 'Platelet count',log=F)
tplot
```



\newpage
## INR

##### Boxplot over time

```{r, fig.width=8}

tplot = boxplot_over_time(inr, 'INR')
tplot
```


##### Plot of means and 95% confidence intervals for the means over time

```{r, fig.width=8}

tplot = means_over_time(inr, 'INR', log=F)
tplot
```



\newpage
## APTT

Values over 200 have been excluded.

##### Boxplot over time

```{r, fig.width=8}

tplot = boxplot_over_time(aptt, 'APTT')
tplot
```

##### Plot of means and 95% confidence intervals for the means over time

```{r, fig.width=8}

tplot = means_over_time(aptt, 'APTT', log=T)
tplot
```




\newpage
## D-dimer

##### Boxplot over time

Values > 70,000 have been excluded

```{r, fig.width=8}


tplot = boxplot_over_time(d_dimer, 'D-dimer')

tplot
```

##### Plot of means and 95% confidence intervals for the means over time

```{r, fig.width=8}


tplot = means_over_time(d_dimer, 'D-dimer', log=T)
tplot
```




\newpage
## IL-6

##### Boxplot over time

Values > 10,000 have been excluded

```{r, fig.width=8}

tplot = boxplot_over_time(il_6, 'IL-6')
tplot
```

##### Plot of means and 95% confidence intervals for the means over time

```{r, fig.width=8}

tplot = means_over_time(il_6, 'IL-6', log=T)
tplot
```



